---
title: "Maternalhealthcombined"
author: Maha Raya, Yamila Emanuel, Evie Starr, Andrea Smith, Desire Omotoye, Unsion
  Park
date: "2025-04-23"
output: html_document
---
#Load packages
```{r loading packages, include = FALSE}
library(haven)
library(ggplot2)
library(ggstats)
```
```{r Import Data, include = FALSE}
maternalhealth <- read.csv(file = "Dataset_maternal_mental_health_infant_sleep.csv")
str(maternalhealth)
```

#Descriptive statistics 
```{r Catergorical Function, include = FALSE}
cat.desc <- function(x)list(
  Frequency = table(x),
  Proportion = (table(x)/length(x[!is.na(x)]))
)
```


```{r Refined Dispersion Function, include = FALSE}
moDisp = function(x, na.rm = TRUE) {
  Range = function(x) {
    max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
  } 
  if (na.rm) {
    x <- x[!is.na(x)]
  }
  
  moDis <- c("Min" = min(x), 
             "Max" = max(x), 
             "Range" = Range(x), 
             "SD" = sd(x), 
             "Variance" = var(x))
             
             return(moDis)
           
}
```


```{r Refined Central Tendency Function, include = FALSE}
moCent <- function(x, na.rm = TRUE) {
    if (na.rm) {
      x <- x[!is.na(x)]
    }
  
  moa <- c(
    "Mean" = mean(x, na.rm = na.rm),
    "Median" = median(x, na.rm = na.rm)
  )
  
  return(moa)
}
```


```{r Age of mother}
#Age of mother (continuous)
moDisp(maternalhealth$Age)
moCent(maternalhealth$Age)
```

```{r Marital Status}
#Marital Status (categorical)
cat.desc(maternalhealth$Marital_status)
```

```{r How baby fell asleep}
#how_falling_asleep_bb1 (categorical)
cat.desc(maternalhealth$how_falling_asleep_bb1)
```

```{r Age of Infant}
#Age of infant (continuous)
moDisp(maternalhealth$Age_bb)
moCent(maternalhealth$Age_bb)
```

#Visualizations
```{r Histogram of one continuous variable with a normal curve}
# Remove rows with NA values in the Age column using subset
maternalhealth_clean <- subset(maternalhealth, Age != "NA")

#simple histogram
SimpAgeHist <- ggplot(maternalhealth_clean, aes(x = Age)) +
  geom_histogram()
SimpAgeHist
#add aesthetics
MotherAgeHist <- ggplot(maternalhealth_clean, aes(x = Age)) +
  geom_histogram(aes(x = Age,
                     after_stat(density)), 
                 alpha = 0.7, 
                 binwidth = 1,
                color="pink" )
MotherAgeHist
#add normal curve
MotherAgeHist2 <- MotherAgeHist + 
  stat_function(fun = dnorm, 
                args = list(mean = mean(maternalhealth$Age, na.rm= TRUE), 
                            sd = sd(maternalhealth$Age, na.rm= TRUE)), 
                col = "purple", lwd = 0.5, lty = "dashed") 
MotherAgeHist2
```
This histogram shows the distribution of the age of the mothers in the dataset. The distribution is approximately normal with a slight right skew. The median age appears to be 30 years old with most individuals ages being concentrated around 25-35 years old. The slight right skew suggests that a signifcant portion of the mothers in the study are 30+ years old.

```{r box plot of a continuous variable }
# 1. Calculate Total Scores 
maternalhealth_df2 <- maternalhealth

maternalhealth_df2$CBTS_M_total <- rowSums(maternalhealth[, c("CBTS_M_3", "CBTS_M_4", "CBTS_M_5", "CBTS_M_6", "CBTS_M_7", "CBTS_M_8", "CBTS_M_9", "CBTS_M_10", "CBTS_M_11", "CBTS_M_12", "CBTS_13", "CBTS_14", "CBTS_15", "CBTS_16", "CBTS_17", "CBTS_18", "CBTS_19", "CBTS_20", "CBTS_21", "CBTS_22")], na.rm = TRUE)
maternalhealth_df2$EPDS_total <- rowSums(maternalhealth[, c("EPDS_1", "EPDS_2", "EPDS_3", "EPDS_4", "EPDS_5", "EPDS_6", "EPDS_7", "EPDS_8", "EPDS_9", "EPDS_10")], na.rm = TRUE)
maternalhealth_df2$HADS_total <- rowSums(maternalhealth[, c("HADS_1", "HADS_3", "HADS_5", "HADS_7", "HADS_9", "HADS_11", "HADS_13")], na.rm = TRUE)

# 2. Create Scale Categories
maternalhealth_df2$CBTS_M_category <- factor(ifelse(maternalhealth_df2$CBTS_M_total <= 20, "Low",  
                             ifelse(maternalhealth_df2$CBTS_M_total <= 40, "Moderate", "High")),  
                             levels = c("Low", "Moderate", "High"))

maternalhealth_df2$EPDS_category <- factor(ifelse(maternalhealth_df2$EPDS_total <= 10, "Low",  
                             ifelse(maternalhealth_df2$EPDS_total <= 20, "Moderate", "High")),  
                             levels = c("Low", "Moderate", "High"))

maternalhealth_df2$HADS_category <- factor(ifelse(maternalhealth_df2$HADS_total <= 7, "Low",  
                             ifelse(maternalhealth_df2$HADS_total <= 14, "Moderate", "High")),  
                             levels = c("Low", "Moderate", "High"))

maternalhealth_df2$night_awakening_number_bb1 <- as.numeric(maternalhealth_df2$night_awakening_number_bb1)

# 3.  Reshape

cbts_df <- data.frame(
  night_awakening_number_bb1 = maternalhealth_df2$night_awakening_number_bb1,
  Category = maternalhealth_df2$CBTS_M_category,
  Scale = "CBTS_M"
)

epds_df <- data.frame(
  night_awakening_number_bb1 = maternalhealth_df2$night_awakening_number_bb1,
  Category = maternalhealth_df2$EPDS_category,
  Scale = "EPDS"
)

hads_df <- data.frame(
  night_awakening_number_bb1 = maternalhealth_df2$night_awakening_number_bb1,
  Category = maternalhealth_df2$HADS_category,
  Scale = "HADS"
)

# Combining  the data frames
df_long <- rbind(cbts_df, epds_df, hads_df)

# 4. Faceted Boxplot ( base R)
par(mfrow = c(1, 3), mar = c(5, 4, 4, 2)) # Arrange plots in 1x3 grid, adjusting margins

scales <- unique(df_long$Scale)

#  Footnotes for each scale definitions
footnotes <- list(
  "CBTS_M" = "CBTS: Low (<=20), Moderate (<=40), High (>40)",
  "EPDS" = "EPDS: Low (<=10), Moderate (<=20), High (>20)",
  "HADS" = "HADS: Low (<=7), Moderate (<=14), High (>14)"
)

# Creating  separate  plots
for (scale_name in scales) {
  scale_data <- df_long[df_long$Scale == scale_name,]
  
  boxplot(scale_data$night_awakening_number_bb1 ~ scale_data$Category,
          main = paste("Scale:", scale_name),  
          xlab = "Score Severity",  
          ylab = "Number of Night Awakenings",  
          cex.axis = 0.8,  # Make x-axis labels smaller to avoid overlap
          border = "black")  # Border color to make boxes visible

  mtext(footnotes[[scale_name]], side = 1, line = 4, cex = 0.5, adj = 0.5)  
}

# main title)
mtext("Infant Night Awakenings by Postpartum Mental Health Scores", side = 3, line = -1, cex = 0.75, font = 2, outer = TRUE)


par(mfrow = c(1, 1)) # Reset plotting parameters

```
This faceted boxplot displays the distribution of the number of times infants aged 0-12 months wake up in the middle of the night according to mothers’ postpartum mental health scores. According to research, the City Birth Trauma Scale (CBTS) ranges from 0-60 with the higher end of the scale indicating greater severity; the Edinburgh Postnatal Depression Scale (EDPS) ranges from 0-40; the Hospital Anxiety and Depression Scale ranges from 0-21. This visualization strongly suggests that infants of mothers who score high in any of these three postpartum mental health surveys tend to wake up more often than infants of the low and moderate scoring severity groups (Sandoz et al, 2021). 
source:Sandoz, Vania & Hingray, Coraline & Ravenscroft/Stuijfzand, Suzannah & Lacroix, Alain & El-Hage, Wissam & Horsch, Antje. (2021). Measurement and Conceptualization of Maternal PTSD Following Childbirth: Psychometric Properties of the City Birth Trauma Scale—French Version (City BiTS-F). Psychological trauma : theory, research, practice and policy. 14. 10.1037/tra0001068.  

```{r pie chart for a categorical variable}
#frequency info
table(maternalhealth$Age_bb)
# Frequency table for marital status saved as new data frame
babyAge <- data.frame(Responses = c("3-6mo", "6-9mo", "9-12 mo"), 
                            "freq" = c(147, 133, 130))

# Generate plot
agePie <- ggplot(babyAge, aes(x = "", 
                                  y = freq, 
                                  fill = factor(Responses))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar(theta = "y") +
  labs(title = "Age Ranges of Participant Babies", 
       caption = "N = 410", 
       fill = "Responses") +
  geom_text(aes(label = paste(round(freq / sum(freq) * 100, 1), "%"), 
                x = 1.7), 
            position = position_stack(vjust = 0.5)) +
    scale_fill_brewer(palette = "Spectral") +
  theme_void()
agePie
```
 There was a pretty even distribution of infant age ranges within the sample population. By a margin of 3.5% or 14 babies, the highest frequency of infants were between 3 and 6 months of age at the time the study was conducted. The lowest frequency of babies were between 9 and 12 months.

```{r bar chart for a categorical variable}
#exclude na values, include = FALSE
clean_barchart <- subset(maternalhealth, Marital_status != "NA")
#making barchart, include = FALSE
barmarital <- ggplot(clean_barchart, aes(x = factor(Marital_status))) +  
  geom_bar(aes(fill = factor(Marital_status)), alpha = 0.8, width = 0.8) + 
  geom_text(aes(y = after_stat(count), label = paste("n = ", after_stat(count))), stat = "count", vjust = -0.5) +
  geom_text(stat = "prop", vjust = 1.5) +
  labs(title = "Marital Status of Participant Mothers", 
       caption = "N = 1047882") + 
  xlab("Marital Status") + 
  ylab("Count") +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "Set2"), 
                    labels = c("Single", "In a relationship", "Separated, divorced or widow", "Other")) +  # Custom legend labels
  scale_x_discrete(breaks = c(1, 2, 3, 6), labels = c("Single", "In a relationship", "Separated, divorced or widow", "Other")) +  # Map labels to breaks
  theme_minimal() +
  theme(legend.position = "none") #exclude legend

barmarital
```
The bar chart presents the distribution of participant mothers based on their marital status. The majority are those in a relationship, making up 94.4% of the sample (n = 387). Smaller categories include single mothers (3.4%, n = 14), those who are separated, divorced, or widowed (1.7%, n = 7), and those categorized as "Other" (0.5%, n = 2). This imbalance reveals that the majority of participants are in committed relationships, while the remaining categories represent a much smaller portion.

```{r scatterplot showing the relationship between the two continuous variables}
# Remove rows with NA values in the Age column using subset
maternalhealth_first_filtered <- subset(maternalhealth, Age != "NA")

#Basic plot Using variables: Age(age of mother) & night_awakening_number_bb1(Night awakenings)
scatterMaternalHealthNight <- ggplot(maternalhealth_first_filtered, aes(x = Age, y = night_awakening_number_bb1)) +
  geom_point() +
  # stat_smooth is the stat equivalent of geom_smooth, "lm" is "linear model" (default is "loess")
  stat_smooth(formula = y ~ x, method = "lm") +
  labs(title = "Mother's Age with Night Awakening's in First Year",
       caption = "N = 410", 
       color = "Age of Babies") +
  xlab("Age of Mother in years") +
  ylab("Night Awakening's") +
  scale_color_brewer(palette = "Paired")

scatterMaternalHealthNight 

#------Detailed plot with Facets using variables: Age(age of mother) & night_awakening_number_bb1(number of times baby wakes up at night) by Age_bb(age of baby in months)--------

# Remove rows with NA values in the Age_bb column using subset
maternalhealth_filtered <- subset(maternalhealth, Age_bb != "NA")
# Creation of Scatterplot
scatterMaternalHealthFacetFinal <- ggplot(maternalhealth_filtered, aes(x = Age, y = night_awakening_number_bb1)) +
  geom_point(aes(color = factor(Age_bb)), 
             shape = 18, alpha = 0.7) +
  stat_smooth(color = "purple", 
              formula = y ~ x,
              method = "loess", #helps handle outliers
              na.rm = TRUE, 
              se = FALSE, 
              linewidth = 0.5, 
              linetype = "dashed") +
  facet_wrap(vars(Age_bb)) +
  labs(title = "Mother's Age with Night Awakening's in First Year by Age of Babies",
       caption = "N = 410", 
       color = "Age of Babies") +
  xlab("Age of Mother in years") +
  ylab("Night Awakening's") +
  scale_color_brewer(palette = "Paired", labels = c("3-6Mo.", "6-9Mo.", "9-12Mo.")) + # Adds custom legend labels)
  theme_minimal() +
  theme(strip.text = element_blank())  # Remove facet labels (numbers)

scatterMaternalHealthFacetFinal
```

 This Scatterplot shows the correlation strength of the mothers age in years ranging between 18 - 100(Age), and the amount of times the infant woke up in the night during the first year(night_awakening_number_bb1); Data is grouped by ordinal variable Age_bb: Age of the baby in months. The regression line in 3-6 months shows a moderate positive correlation in awakenings during the ages 25-32, with an uptick as age increases; Indicating post prime-childbearing years, night awakenings can increase in infants. The regression in 6-9 months shows a moderate negative correlation, with a significant drop in awakenings within the ‘high-risk’ pregnancy age(35+); This can signify that as the mother ages, the amount of awakenings decrease in this infant age range. The 9-12 months regression line presents a strong positive correlation between mother’s age and awakenings, with a significant trend from an estimated occurrence of 1-2 times to upwards of 3; Non-linear relationship shows some data points deviating significantly from overall trend.
 
#Hypothesis
-Research Question
How does a baby's age influence their preference of being put to sleep by parents ?
-Question 1
Null hypothesis (H0: μ = 0.8): The mean number of infant night awakenings (age range: 3 - 6 months)  is equal to 0.8. 
Alternate hypothesis (H1:μ ≠ 0.8): The mean number of infant night awakenings  (age range: 3 - 6 months) is not equal to 0.8. 

The purpose of this test is to evaluate whether the observed sample mean of infant night awakenings deviates significantly from the hypothesized mean of 0.8 awakenings per night. We can see if the typical number of night awakenings aligns with expectations based on previous studies.
Reference: Galland, B. C., Taylor, B. J., Elder, D. E., & Herbison, P. (2012). Normal sleep patterns in infants and children: a systematic review of observational studies. Sleep medicine reviews, 16(3), 213–222. https://doi.org/10.1016/j.smrv.2011.06.001

-Question 2:t-test
t-test Null hypothesis (H0): There is a relationship between a mother’s marital status (Marital_status)  and her baby’s sleep pattern (night_awakening_number_bb1). 
µ1=µ2
t-test Alternate hypothesis (H1): There is no relationship between a mother’s marital status (Marital_status) and her baby’s sleep pattern (night_awakening_number_bb1). 
µ1≠µ2
To assess whether our data meets the assumptions for conducting an independent sample t-test, we will first subset the data based on the categorical grouping variable, Marital_status ( 1 = single,  2 =  in a relationship).  We will examine the distribution of the continuous variable (number of infant night awakenings) for each group using  Q-Q plot and histogram to visually assess normality. Afterwards we will formally check the variance equality within each group using the variance test (var.test).
```{r t-test}
maternalhealth$Marital_status <- factor(maternalhealth$Marital_status, 
                            levels = c("1", 
                                       "2"     ), 
                            ordered = TRUE)
#consider re-coding levels from 1 and 2 to 0 and 1. 
#Further re-coding would not align with the values in the code book.
str(maternalhealth$Marital_status)
2
t.test(night_awakening_number_bb1 ~ Marital_status, data = maternalhealth, alternative = "two.sided",
       var.equal = FALSE)

```

-Question 3: Chi-square
Chi-square Null hypothesis (H0): There is no relationship between the age of the baby (Age_bb)  and the method in which the baby falls asleep (how_falling_asleep_bb1). 
Chi-square Alternate hypothesis (H1): There is a relationship between the age of the baby (Age_bb)  and the method in which the baby falls asleep (how_falling_asleep_bb1).
```{r chi-square}
pregSleep <- subset(maternalhealth, select = c("Age_bb", "how_falling_asleep_bb1"))
pregSleep$Age_bb <- as.character(pregSleep$Age_bb)
pregSleep$how_falling_asleep_bb1<-as.character(pregSleep$how_falling_asleep_bb1)

## Assumptions check (Meets the assumption)
table(pregSleep$Age_bb, pregSleep$how_falling_asleep_bb1)
# 15 total cells; 8 are under 5; make sure 80% are over at least 5 and over
#12/12 = 100% 

chiSq <- chisq.test(pregSleep$Age_bb, y = pregSleep$how_falling_asleep_bb1)
chiSq
```
-Question 4: Variables selected for multiple linear regression analysis
Variables from the dataset selected for multiple linear regression analysis, and the outcome with text that describes how we will assess if our data meet assumptions for the statistical test: 
*Predictor: EPDS scale
*Predictor: Duration of Sleep
*Outcome: Infant night awakening
```{r standardization}
maternalhealth_df1 <- maternalhealth
# Step 1: Convert to decimal hours
maternalhealth_df1$Sleep_night_duration_bb1 <- sapply(strsplit(as.character(maternalhealth$Sleep_night_duration_bb1), ":"), function(x) {
  as.numeric(x[1]) + as.numeric(x[2])/60
})

maternalhealth_df1$EPDS_total <- rowSums(maternalhealth[, c("EPDS_1", "EPDS_2", "EPDS_3", "EPDS_4", "EPDS_5", "EPDS_6", "EPDS_7", "EPDS_8", "EPDS_9", "EPDS_10")], na.rm = TRUE)
```
```{r examine variables of interest}
maternalhealth_interest <- maternalhealth_df1[, c("EPDS_total", "night_awakening_number_bb1", "Sleep_night_duration_bb1")]


# Use the summary command to examine variables
summary(maternalhealth_interest)
```
```{r handle NAs}
sum(!is.na(maternalhealth_interest$EPDS_total))
# 410
sum(is.na(maternalhealth_interest$EPDS_total))
# 1047882


maternalhealth_interest$Sleep_night_duration_bb1[maternalhealth_interest$Sleep_night_duration_bb1 == 100.65] <- NA
# This is re-coded as NA because it was originally in the dataset as “99:99”, which indicates that it is a missing value and not an extreme outlier


clean_maternal_lm_data <- na.omit(maternalhealth_interest[, c("night_awakening_number_bb1", "EPDS_total", "Sleep_night_duration_bb1")])
```
```{r variables of interest recheck}


summary(clean_maternal_lm_data)
```
**We will determine if our data meets the assumptions by running the following to assess the distribution of our data…**
```{r regression}
model1 <- lm(night_awakening_number_bb1 ~ EPDS_total + Sleep_night_duration_bb1, data = clean_maternal_lm_data)
summary(model1)
# Print coefficient 95% confidence intervals
confint(model1)
# Diagnostics 
plot(model1, las = 1)
```
Based on our results, a number of assumptions are violated. Our data is heteroscedastic and has some outliers as seen with points labeled 240, 237, and 280 in the Q-Q Residuals plot as well as the Cooks *d* and Leverage. This is a limitation we faced here as these influential observations may skew our results and are an indication of non-normal residuals . It is also important to note here that Sleep_night_duration_bb1, which is one of our predictors, varies greatly across baby age groups; this may explain the heteroscedasticity of our data. Our dataset covers babies from the ages of 3-12 months old.
```{r scatter-2}
# Create plot of regression line for age 
EPDS_reg_plot <- ggplot(clean_maternal_lm_data, aes(x = EPDS_total, y = night_awakening_number_bb1)) +
  geom_point(na.rm = TRUE, 
             shape = 12, 
             alpha = 0.8, 
             color = "purple4", 
             # Adding jitter, as number of days is an integer (shows concentration of points)
             position = position_jitter(width = 0.1)) +
  geom_smooth(method = "lm", 
              color = "blue2", 
              linewidth = 0.3, 
              se = FALSE, 
              na.rm = TRUE) +
  labs(title = "Night Awakenings vs EPDS Total") +
  xlab("Edinburgh Postnatal Depression Scale (EDPS)") +
  ylab("Night Awakenings") +
  theme_minimal()


EPDS_reg_plot


Sleep_reg_plot <- ggplot(clean_maternal_lm_data, aes(x = Sleep_night_duration_bb1, y = night_awakening_number_bb1)) +
  geom_point(na.rm = TRUE, 
             shape = 12, 
             alpha = 0.8, 
             color = "purple4", 
             # Adding jitter, as number of days is an integer (shows concentration of points)
             position = position_jitter(width = 0.1)) +
  geom_smooth(method = "lm", 
              color = "blue2", 
              linewidth = 0.3, 
              se = FALSE, 
              na.rm = TRUE) +
  labs(title = "Night Awakenings vs Sleep Duration") +
  xlab("Sleep Duration at Night (hrs)") +
  ylab("Night Awakenings") +
  theme_minimal()
Sleep_reg_plot
```
*Note:* Noise was added to scatterplot to display clustering of points and points are not absolute reflections of the data. 
The EPDS vs. Night Awakenings scatterplot shows a slight upward trend, which is reflected in the positive slope of the fitted regression line; The Sleep Duration vs. Night Awakenings plot shows a slight downward trend, indicating a negative slope of the fitted regression line. Neither of the plots above demonstrate a strong linear relationship because the spread is weak. The high variance in night awakenings across all values of both predictors suggests a nonlinear relationship.
```{r VIF function}
vif_func <- function(model1) {
  # Extract the model matrix
  X <- model.matrix(model1)[, -1]
  
  # Calculate VIF for each predictor
  vif_values <- sapply(1:ncol(X), function(i) {
    
    # Fit the model
    reg <- lm(X[, i] ~X[, -i])
    
    # Calculate VIF
    1 / (1-summary(reg)$r.squared)
  })
  
  # Name the VIF values with predictor names
  names(vif_values) <- colnames(X)
  
  # Print results
  return(vif_values)
}
```
```{r check VIF}
vif_func(model1)
```
The VIF for both predictors is 1.040626, which is well below 3. This indicates that our two predictors are not highly correlated with each other, and there's no concern about multicollinearity in our model. Our assumption of no multicollinearity is checked and we can confidently include both variables in a regression model.
```{r evaluate outliers}
# By Cook's d
cooksd <- cooks.distance(model1)


# Plot by Cooks d
plot(cooksd, pch = "", cex = 2, main = "Influential Obs by Cooks d")
abline(h = 4*mean(cooksd, na.rm = TRUE), col = "red")
text(x = 1:length(cooksd) + 1, y = cooksd, 
     labels = ifelse(cooksd > 4*mean(cooksd, na.rm = TRUE), 
                     names(cooksd), ""), col = "red")


# Look in dataset
influential <- as.numeric(names(cooksd)[(cooksd > 4*mean(cooksd, na.rm = T))])


list(maternalhealth_interest[influential, ])
```
We used Cook's distance to inspect and evaluate the outliers. After inspecting the outliers here, we have concluded that they are not extreme or a cause for concern. Initially, there was *one* outlier that was found that was was in the original dataset as "99:99", which indicates missing data, rather than a true time value. We re-coded it as NA and omitted it with the other NAs.
```{r robust standard errors}
# Load packages
library("lmtest")
#install.packages("sandwich")
library("sandwich")


# Re-run model
model1_rb <- coeftest(model1, 
                     vcov = vcovHC)


model1_rb
# Compare to original
summary(model1)
```
Because we observed heteroscedasticity in our residuals, we have executed robust standard errors to account for this violation of assumption for multiple linear regression. From the results, we found that when accounting for heteroscedasticity, the intercept is highly significant, with a p-value of 2.86e-13, the EPDS_total is *not* significant, with a p-value of 0.1717, and the Sleep_night_duration_bb1 is also highly significant, with a p-value of 2.26e-08. The negative coefficient indicates that longer sleep duration is associated with fewer night awakenings. The overall F-statistic is 3.287e-09, and indicates that the independent variables in our model collectively explain a statistically significant amount of the variance in the dependent variable (night_awakening_number_bb1).

-Question 5: Variables selected for logistic regression analysis
Variables from the dataset selected for logistic regression analysis, and the outcome:
*Predictor: Marital Status | Marital_status 
*Predictor: Postnatal Depression Score | EPDS_7 | “Difficulty sleeping due to unhappiness"
Reference for EPDS Scale - item 7: https://www.cope.org.au/health-professionals/health-professionals-3/calculating-score-epds/
*Outcome: Infant Sleep Duration | Sleep_night_duration_bb1 | 0 is >= 9 hours (good sleep) vs 1 is < 9 hours (short sleep)
Reference for average sleep duration parameter at 9 hours: https://pmc.ncbi.nlm.nih.gov/articles/PMC8627435/

#Results and interpretations
-Question 1
```{r A test of normality, including measures of skewness, comparing the mean to the median, histogram with normal curve overlaid, normal q-q plot, and significance test for normality for the continuous variable used in the following two tests }

shapiro.test(maternalhealth$night_awakening_number_bb1)
```	

Shapiro-Wilk normality test
data:  maternalhealth$night_awakening_number_bb1
 W = 0.8066, p-value < 2.2e-16

Low p-value indicates non-normality of distribution because the Shapiro-Wilk test null hypothesis is that data are not normally distributed. 

Skewness: 1.731061 (had to set na.rm = TRUE)

```{r Measures of skewness}
#measures of skewness
Skew <- function(x, na.rm = TRUE){
  if(na.rm) x <- x[!is.na(x)]
  n <- length(x)
  sum((x - mean(x))^3)/(n - 2)/var(x)^(3/2)
}

Skew(maternalhealth$night_awakening_number_bb1)

#Histogram
histAwake <- ggplot(maternalhealth, aes(night_awakening_number_bb1)) + 
  geom_histogram(aes(x = night_awakening_number_bb1, after_stat(density)), fill = 'lightgreen', col = 'black') + 
 geom_density() + 
                  labs(title = "Distribution of Night Awakenings", 
                       caption = "N = 1048292") + 
                  xlab("Number of Night Awakenings") + 
                  ylab("Density") + 
                  theme_minimal()

```
Compare mean to median:
Mean = 1.439024 (had to set na.rm = TRUE)
Median = 1 (had to set na.rm = TRUE) 

```{r Histogram with normal curve overlay: } 
histAwake
```

```{r QQ Plot}
qqnorm(maternalhealth$night_awakening_number_bb1, pch = 1, frame = FALSE)
qqline(maternalhealth$night_awakening_number_bb1, col = "blue", lwd = 2)

```
The qq-plot is nonlinear, indicating a skewed distribution of the datapoints. This aligns with the histogram, which shows the data are left-skewed toward fewer night awakenings. Additionally, this histogram line of fit highlights that the data are not normally distributed. 

Significance test for normality-
Spearman's rank correlation rho:
data:  sleepmat$night_awakening_number_bb1 and sleepmat$Marital_status
S = 11875680, p-value = 0.4942
alternative hypothesis: true rho is not equal to 0
sample estimates:
        rho 
-0.03385764
 → weak negative correlation between marital status and number of night awakenings. 


-Question 2
```{r A test for H0: μ =/</>, of your hypothesized value, including the 95%CI}
newDf <- subset(maternalhealth, select = c("Age_bb"))
agerange <- subset(newDf, select = c(1))
t.test(agerange, mu = 0.8,  alternative = "two.sided")
```

Interpretation of results: The results of the one-sample t-test reveal that the true mean of times infants aged from 3-6 months wake up at night is not equal to  0.08. ($\bar{x} = 1.96 ) *p* < 0.05).  The 95% confidence interval for the true mean lies between 1.88 and 2.04.

-Question 3
```{r An independent samples or paired samples t-test (as appropriate), including the 95%CI}
maternalhealth$Marital_status1 <- factor(maternalhealth$Marital_status, levels = c(1, 2), labels = c("Single", "In a Relationship"))
t_test_result <- t.test(how_falling_asleep_bb1 ~ Marital_status1, data = maternalhealth, alternative = "two.sided", var.equal = FALSE)
print(t_test_result)
```
Interpretation: The results from the two-sample t-test indicate a p-value of 0.471 which is greater than 0.05, thus we can conclude that there is no statistically significant difference between  the mean number of times a baby wakes up during the night between mothers who are single or those in a relationship. We are 95% confident that the true difference in the mean number of nighttime awakenings for these two marital status is between -0.67 and 1.38. Also, since the interval includes 0 we can also conclude that there is still no  significant difference in night awakenings based on marital status.

-Question 4
```{r A contingency table with a Chi-Square significant test for H0: independent between categorical variables. Include an odds ratio and relative risk for one comparison of categorical variables}
table(maternalhealth$Age_bb, maternalhealth$how_falling_asleep_bb1)
chiSq <- chisq.test(maternalhealth$Age_bb, y = maternalhealth$how_falling_asleep_bb1)
chiSq
chiSq$observed
chiSq$expected
chiSq$residuals
```
Interpretation: We reject the null hypothesis that there is not an association between the method in which the baby falls asleep and the age of the baby ($\chi^2$(8) = 19.09, *p* < .05). Specifically, those who are 3 to 6 months old are more likely to fall asleep while being fed, and less likely to fall asleep in the crib alone. Babies who are 6 to 9 months are more likely to fall asleep alone in the crib than expected. Most interestingly, babies who are 9 to 12 months are more likely to fall asleep in the crib with parental presence as opposed to being alone in the crib. The relative risk of a 3 to 6 month year old infant falling asleep while being fed is 1.87. Thus, we can  conclude that 3 to 6 month olds are 1.87 times more likely to fall asleep while being fed as compared to all other methods of putting a baby to sleep.

-Question 5
```{r A scatterplot of two continuous variables, and a linear correlation test for H0: π = 0}
scatterCorr <- ggplot(maternalhealth, aes(x =Age, y = night_awakening_number_bb1)) +
  geom_point(na.rm = TRUE, 
             shape = 13, 
             alpha = 0.8, 
             color = "purple2") +
  labs(title = "Baby's Night awakening by age of mother") +
  xlab("Age in years") + 
  ylab("Number of night awakening") + 
  theme_minimal()
scatterCorr

#Linear Correlation test
# Subset variables of interest
nightage <- subset(maternalhealth, select = c("night_awakening_number_bb1", "Age"))
# Data cleanup - age
summary(nightage$Age)
class(nightage$Age)

# Turn "Refused" to NAs; coerce as numeric
nightage$Age[nightage$Age == "Refused"] <- NA
nightage$Age <- as.numeric(nightage$Age)

# Data cleanup - Night awakening
summary(nightage$night_awakening_number_bb1)
class(nightage$night_awakening_number_bb1)

# Turn "Refused" to NAs; coerce as numeric
# Assumptions checks for normality 
shapiro.test(nightage$Age)

qqnorm(nightage$Age, pch = 1, frame = FALSE)
qqline(nightage$Age, col = "purple", lwd = 2)

shapiro.test(nightage$night_awakening_number_bb1)

qqnorm(nightage$night_awakening_number_bb1, pch = 1, frame = FALSE)
qqline(nightage$night_awakening_number_bb1, col = "pink", lwd = 2)

# Run Pearson's R, store as new object
nightagecorr <- cor.test(nightage$Age ,nightage$night_awakening_number_bb1,
                       method = "pearson", 
                       alternative = "two.sided", 
                       na.action = na.exclude,
                       exact = TRUE)

nightagecorr 
nightagecorr $conf.int

```
Interpretation: The scatter plot of baby's night awakenings by mother's age shows widely dispersed points with no clear linear pattern, suggesting minimal correlation between maternal age and nighttime wakings. While some outliers exist showing higher awakening frequencies (7-10 times) among mothers in their early 30s, most data points cluster between 0-5 awakenings across all age groups. The Q-Q plots indicate that while maternal age follows a relatively normal distribution, the night awakening variable shows significant deviation from normality with a step-like pattern and many zero values, suggesting a discrete, right-skewed distribution. This non-normality in the dependent variable may impact the validity of the Pearson correlation test used in the analysis.

-Question 6
```{r A scatterplot with estimated simple(ordinary) regression line}
# Subset variables of interest
nightage <- subset(maternalhealth, select = c("night_awakening_number_bb1", "Age"))
# Data cleanup - Age
summary(nightage$Age)
class(nightage$Age)




# Turn "Refused" to NAs; coerce as numeric
nightage$Age[nightage$Age == "Refused"] <- NA
nightage$Age <- as.numeric(nightage$Age)




# Data cleanup - Night awakening
summary(nightage$night_awakening_number_bb1)
class(nightage$night_awakening_number_bb1)
```
```{r Night/Age Ordinary Regression Scatterplot w/ lm line}
# Scatter for Age and Night Awakenings
scatterLm <- ggplot(nightage, aes(x = Age, y = night_awakening_number_bb1)) +
  geom_point(na.rm = TRUE, 
             shape = 13, 
             alpha = 0.8, 
             color = "purple") +
  geom_smooth(method = "lm", 
              color = "red",
              formula = y ~ x,
              linewidth = 0.3, 
              se = FALSE, 
              na.rm = TRUE) +
  labs(title = "Age of Mother by The Number of Night Awakenings") +
  xlab("Age in years") + 
  ylab("Night Awakenings") + 
  theme_minimal()

scatterLm
```
With the addition of the fit line, it does not move through points in our actual data; The line is relatively flat, indicating that the relationship between the age of the mother and number of infant night awakenings is not specifically strong

-Question 7
```{r A simple linear regression analysis, including a residual plot from the above linear regression analysis}
# Remove rows with NA values in the Age column using subset
maternalhealth_first_filtered <- subset(maternalhealth, Age != "NA")


# To demonstrate the link between correlation and OLS Ord. Regression; same nightage
nightageLM <- lm(night_awakening_number_bb1 ~ scale(Age), data = nightage, 
               na.action = na.exclude)


summary(nightageLM)


# Diagnostic plots 
plot(nightageLM, las = 1)
```
Interpretation: We fail to reject the null hypothesis that there is no association between age and number of night awakenings (F(1, 408) = 0.41, p = .524 | β = 0.05, t = 0.64, p = .524, R^2 = 0.001, Adjusted R^2 = -0.001). There is no significant association between age and night awakenings, such that one standard deviation increase in age predicts a 0.05 increase in night awakenings. However, we violate assumptions of normality of residuals and outliers are present; The residual distribution presents as having a right tail, indicating positive skewness. Thus, these results should be taken with caution.

-Question 8
```{r A multiple regression analysis, including variable selection,A scatterplot with significant values from the multiple regression analysis (1pt)}
maternalhealth_df1 <- maternalhealth
# Step 1: Convert to decimal hours
maternalhealth_df1$Sleep_night_duration_bb1 <- sapply(strsplit(as.character(maternalhealth$Sleep_night_duration_bb1), ":"), function(x) {
  as.numeric(x[1]) + as.numeric(x[2])/60
})

maternalhealth_df1$EPDS_total <- rowSums(maternalhealth[, c("EPDS_1", "EPDS_2", "EPDS_3", "EPDS_4", "EPDS_5", "EPDS_6", "EPDS_7", "EPDS_8", "EPDS_9", "EPDS_10")], na.rm = TRUE)
```

```{r examine variables of interest}
maternalhealth_interest <- maternalhealth_df1[, c("EPDS_total", "night_awakening_number_bb1", "Sleep_night_duration_bb1")]


# Use the summary command to examine variables
summary(maternalhealth_interest)
```
```{r handle NAs}
sum(!is.na(maternalhealth_interest$EPDS_total))
# 410
sum(is.na(maternalhealth_interest$EPDS_total))
# 1047882


maternalhealth_interest$Sleep_night_duration_bb1[maternalhealth_interest$Sleep_night_duration_bb1 == 100.65] <- NA
# This is re-coded as NA because it was originally in the dataset as “99:99”, which indicates that it is a missing value and not an extreme outlier


clean_maternal_lm_data <- na.omit(maternalhealth_interest[, c("night_awakening_number_bb1", "EPDS_total", "Sleep_night_duration_bb1")])
```
```{r variables of interest recheck}


summary(clean_maternal_lm_data)
```
**We will determine if our data meets the assumptions by running the following to assess the distribution of our data…**
```{r regression}
model1 <- lm(night_awakening_number_bb1 ~ EPDS_total + Sleep_night_duration_bb1, data = clean_maternal_lm_data)
summary(model1)
# Print coefficient 95% confidence intervals
confint(model1)
# Diagnostics 
plot(model1, las = 1)
```
Based on our results, a number of assumptions are violated. Our data is heteroscedastic and has some outliers as seen with points labeled 240, 237, and 280 in the Q-Q Residuals plot as well as the Cooks *d* and Leverage. This is a limitation we faced here as these influential observations may skew our results and are an indication of non-normal residuals . It is also important to note here that Sleep_night_duration_bb1, which is one of our predictors, varies greatly across baby age groups; this may explain the heteroscedasticity of our data. Our dataset covers babies from the ages of 3-12 months old.
```{r scatterplot}
# Create plot of regression line for age 
EPDS_reg_plot <- ggplot(clean_maternal_lm_data, aes(x = EPDS_total, y = night_awakening_number_bb1)) +
  geom_point(na.rm = TRUE, 
             shape = 12, 
             alpha = 0.8, 
             color = "purple4", 
             # Adding jitter, as number of days is an integer (shows concentration of points)
             position = position_jitter(width = 0.1)) +
  geom_smooth(method = "lm", 
              color = "blue2", 
              linewidth = 0.3, 
              se = FALSE, 
              na.rm = TRUE) +
  labs(title = "Night Awakenings vs EPDS Total") +
  xlab("Edinburgh Postnatal Depression Scale (EDPS)") +
  ylab("Night Awakenings") +
  theme_minimal()


EPDS_reg_plot


Sleep_reg_plot <- ggplot(clean_maternal_lm_data, aes(x = Sleep_night_duration_bb1, y = night_awakening_number_bb1)) +
  geom_point(na.rm = TRUE, 
             shape = 12, 
             alpha = 0.8, 
             color = "purple4", 
             # Adding jitter, as number of days is an integer (shows concentration of points)
             position = position_jitter(width = 0.1)) +
  geom_smooth(method = "lm", 
              color = "blue2", 
              linewidth = 0.3, 
              se = FALSE, 
              na.rm = TRUE) +
  labs(title = "Night Awakenings vs Sleep Duration") +
  xlab("Sleep Duration at Night (hrs)") +
  ylab("Night Awakenings") +
  theme_minimal()
Sleep_reg_plot
```
*Note:* Noise was added to scatterplot to display clustering of points and points are not absolute reflections of the data. 

The EPDS vs. Night Awakenings scatterplot shows a slight upward trend, which is reflected in the positive slope of the fitted regression line; The Sleep Duration vs. Night Awakenings plot shows a slight downward trend, indicating a negative slope of the fitted regression line. The positive slope in the EPDS vs. Night awakenings scatterplot plot tells us that as the score increases, infant night awakenings also increase, although the association is not strong. The trend in the second plot tells us that as infants sleep longer at night, they tend to wake up fewer times during the night; longer the durations of sleep among infants are associated with fewer night awakenings. Neither of the plots above demonstrate a strong linear relationship because the spread is weak. The high variance in night awakenings across all values of both predictors suggests a nonlinear relationship.
```{r VIF function}
vif_func <- function(model1) {
  # Extract the model matrix
  X <- model.matrix(model1)[, -1]
  
  # Calculate VIF for each predictor
  vif_values <- sapply(1:ncol(X), function(i) {
    
    # Fit the model
    reg <- lm(X[, i] ~X[, -i])
    
    # Calculate VIF
    1 / (1-summary(reg)$r.squared)
  })
  
  # Name the VIF values with predictor names
  names(vif_values) <- colnames(X)
  
  # Print results
  return(vif_values)
}
```
```{r check VIF}
vif_func(model1)
```
The VIF for both predictors is 1.040626, which is well below 3. This indicates that our two predictors are not highly correlated with each other, and there's no concern about multicollinearity in our model. Our assumption of no multicollinearity is checked and we can confidently include both variables in a regression model.
```{r evaluate outliers}
# By Cook's d
cooksd <- cooks.distance(model1)


# Plot by Cooks d
plot(cooksd, pch = "", cex = 2, main = "Influential Obs by Cooks d")
abline(h = 4*mean(cooksd, na.rm = TRUE), col = "red")
text(x = 1:length(cooksd) + 1, y = cooksd, 
     labels = ifelse(cooksd > 4*mean(cooksd, na.rm = TRUE), 
                     names(cooksd), ""), col = "red")


# Look in dataset
influential <- as.numeric(names(cooksd)[(cooksd > 4*mean(cooksd, na.rm = T))])


list(maternalhealth_interest[influential, ])
```
We used Cook's distance to inspect and evaluate the outliers. After inspecting the outliers here, we have concluded that they are not extreme or a cause for concern. Initially, there was *one* outlier that was found that was was in the original dataset as "99:99", which indicates missing data, rather than a true time value. We re-coded it as NA and omitted it with the other NAs.
```{r robust standard errors}
# Load packages
library("lmtest")
#install.packages("sandwich")
library("sandwich")


# Re-run model
model1_rb <- coeftest(model1, 
                     vcov = vcovHC)


model1_rb
# Compare to original
summary(model1)
```
Because we observed heteroscedasticity in our residuals, we have executed robust standard errors to account for this violation of assumption for multiple linear regression. From the results, we found that when accounting for heteroscedasticity, the intercept is highly significant, with a p-value of 2.86e-13, the EPDS_total is *not* significant, with a p-value of 0.1717, and the Sleep_night_duration_bb1 is also highly significant, with a p-value of 2.26e-08. The negative coefficient indicates that longer sleep duration is associated with fewer night awakenings. The overall F-statistic is 3.287e-09, and indicates that the independent variables in our model collectively explain a statistically significant amount of the variance in the dependent variable (night_awakening_number_bb1).

-Question 9
```{r A logistic regression analysis }
# Subset marital status, sleep duration of infant, and EPDS_7: "I have been so unhappy that I have had difficulty sleeping"

maternalLog <- subset(maternalhealth, select = c("Sleep_night_duration_bb1", "Marital_status", "EPDS_7"))


summary(maternalLog)
```


```{r logreg data cleanup}
# Convert hh:mm character to numeric hours
maternalLog$sleep_duration_hours <- as.numeric(substr(maternalLog$Sleep_night_duration_bb1, 1, 2)) +
                                    as.numeric(substr(maternalLog$Sleep_night_duration_bb1, 4, 5)) / 60


# Recode sleep_night_duration_bb1 to binary:
# 0 = ≥ 9 hours (good sleep), 1 = < 9 hours (short sleep)
maternalLog$sleep_duration_binary[maternalLog$sleep_duration_hours >= 9] <- 0
maternalLog$sleep_duration_binary[maternalLog$sleep_duration_hours < 9] <- 1




# Clear invalid responses NA
maternalLog <- subset(maternalLog, !is.na(sleep_duration_binary) & !is.na(Marital_status) & !is.na(EPDS_7))


# Factorize Marital_status 
maternalLog$Marital_status <- factor(maternalLog$Marital_status,
                                     levels = c(1, 2, 3, 6),
                                     labels = c("Single", "In a Relationship", "Separated/Divorced/Widow", "Other"))
summary(maternalLog)


# Check against frequencies
table(maternalLog$sleep_duration_binary)
table(maternalLog$Marital_status)
table(maternalLog$EPDS_7)
```


```{r run logistic regression}
maternalLogR <- glm(sleep_duration_binary ~ Marital_status + EPDS_7,
             data = maternalLog, family = "binomial")


# View results summary
summary(maternalLogR)


# Get ORs and 95%CIs
exp(cbind(OR = coef(maternalLogR), confint(maternalLogR)))


# Get pseudo R2
1 - (maternalLogR$deviance / maternalLogR$null.deviance)
```


```{r VIF function}
vif_func <- function(model) {
  # Extract the model matrix
  X <- model.matrix(model)[, -1]
  
  # Calculate VIF for each predictor
  vif_values <- sapply(1:ncol(X), function(i) {
    
    # Fit the model
    reg <- lm(X[, i] ~X[, -i])
    
    # Calculate VIF
    1 / (1-summary(reg)$r.squared)
  })
  
  # Name the VIF values with predictor names
  names(vif_values) <- colnames(X)
  
  # Print results
  return(vif_values)
}
```

```{r check VIF}
vif_func(maternalLogR)
```
None of our variables have VIF > 5, we have likely met the no multicollinearity assumption.

```{r binned residuals for con. variable}
# For EPDS_7, seen as continuous 
performance::binned_residuals(maternalLogR, term = "EPDS_7")
```
The model fit is not great; Only an estimated 50% of data fit within the assumed linear relationship between the predictors and logit of the outcome.


```{r test}
# Check cross-tabulations to see if we have any perfect predictors
table(maternalLog$sleep_duration_binary, maternalLog$EPDS_7)
```

Interpretation: Overall, the model predicted approximately 2.3% of the variance in the probability of experiencing shorter sleep durations (Pseudo R^2 = 0.023; AIC = 306.98). For each one-unit increase of the Mother having experienced unhappiness, was associated with a 1.39 times greater likelihood of reporting shorter sleep duration (95% CI = 1.05, 1.83; $\beta$ = 0.33, z = 2.34, p = .019). Those who were in a relationship had a 0.82 times lower likelihood of experiencing shorter sleep (95% CI = 0.21, 5.46; $\beta$ = -0.19, z = -0.25, p = .806), as compared to Mothers who identified as single. Similarly, Mothers who were separated, divorced, or widowed had 1.22 times the probability of experiencing shorter sleep (95% CI = 0.05, 15.83; $\beta$ = 0.20, z = 0.15, p = .883); Further, those identifying as "Other" increases by 6.94 (95% CI = 0.21, 243.04; $\beta$ = 1.94, z = 1.20, p = .231). There are potential misfit issues, there is evidence to suggest that one ordinal variable - EPDS_7 does not have a linear relationship with the outcome of sleep duration.